### 2017-04-07 paper/oopsla17-submission/evaluation-accuracy.tex
update numbers

\paragraph{Our Classifiers}
We evaluate five classifiers, each trained on the full set of features.
% features: 44 local syntactic features, 176 contextual syntactic
% features, 55 typing features, and a single expression size feature.
% %
% \ES{should explain the make-up of these groups}
%
% We preemptively discard expressions that are not part of the minimal
% type error slice --- we will explain the rationale for this in
% \autoref{sec:feature-utility} --- and thus the final feature count is
% 276.
%
Our classifiers are:
%
\begin{enumerate}
\item A logistic regression trained with a learning rate
  $\eta = 0.001$, an $L_2$ regularization rate $\lambda = 0.001$, and a
  mini-batch size of 200.
\item A decision tree trained with the \ES{HUMA?} algorithm and \ES{HUMA: hyperparams?}.
\item A random forest trained with the \ES{HUMA?} algorithm and 30 estimators \ES{HUMA: hyperparams?}.
\item Two multi-layer perceptron neural networks, both trained with $\eta = 0.001$,
  $\lambda = 0.001$, and a mini-batch size of 200. The first MLP
  contains a single hidden layer of 10 neurons, and the second contains
  a hidden layer of 500 neurons. This allows us to investigate how well
  the MLP can \emph{compress} its model (cf.~\cite{FIXME}). The neurons
  use rectified linear units (ReLU) as their activation function, a
  common practice in modern neural networks.
\end{enumerate}
%
All classifiers were trained for 20 epochs on one dataset
--- \ie they were shown each program 20 times ---
before being evaluated on the other.
%
The logistic regression and MLPs were trained with the \textsc{Adam}
optimizer \citep{Kingma2014-ng}, a variant of stochastic gradient
descent that has been found to converge faster.

>>>>

\paragraph{Our Classifiers}
We evaluate five classifiers, each trained on the full set of features.
% features: 44 local syntactic features, 176 contextual syntactic
% features, 55 typing features, and a single expression size feature.
% %
% \ES{should explain the make-up of these groups}
%
% We preemptively discard expressions that are not part of the minimal
% type error slice --- we will explain the rationale for this in
% \autoref{sec:feature-utility} --- and thus the final feature count is
% 276.
%
Our classifiers are:
%
\begin{description}
\item[\linear] A logistic regression trained with a learning rate
  $\eta = 0.001$, an $L_2$ regularization rate $\lambda = 0.001$, and a
  mini-batch size of 200.
\item[\dectree] A decision tree trained with the \ES{HUMA?} algorithm and \ES{HUMA: hyperparams?}.
\item[\forest] A random forest trained with the \ES{HUMA?} algorithm and 30 estimators \ES{HUMA: hyperparams?}.
\item[\hiddenT and \hiddenFH] Two multi-layer perceptron neural networks, both trained with $\eta = 0.001$,
  $\lambda = 0.001$, and a mini-batch size of 200. The first MLP
  contains a single hidden layer of 10 neurons, and the second contains
  a hidden layer of 500 neurons. This allows us to investigate how well
  the MLP can \emph{compress} its model (cf.~\cite{FIXME}). The neurons
  use rectified linear units (ReLU) as their activation function, a
  common practice in modern neural networks.
\end{description}
%
All classifiers were trained for 20 epochs on one dataset
--- \ie they were shown each program 20 times ---
before being evaluated on the other.
%
The logistic regression and MLPs were trained with the \textsc{Adam}
optimizer \citep{Kingma2014-ng}, a variant of stochastic gradient
descent that has been found to converge faster.

### 2017-04-07 paper/oopsla17-submission/evaluation-utility.tex
update numbers

We tested our hypothesis with the \linear and
%
\hiddenFH\footnote{A layer of 500 neurons is excessive when we have so few
  input features --- we use \hiddenFH for continuity with the
  surrounding sections.}
%
classifiers, cross-validated ($k=10$) over the combined SP14/FA15
dataset. We used a learning rate $\eta=0.001$, $L_2$ regularization rate
$\lambda=0.001$, and mini-batch size of 200. We trained for a single
epoch on feature sets (1) and (2), and for 8 epochs on (3), so that the
total number of training samples would be roughly equal for each feature
set.
%
\lstDeleteShortInline{|} % sigh...
In addition to the accuracy metric from before, we report each
classifier's \emph{recall} --- \ie ``How many true changes can we
remember?'' --- defined as
$$
\frac{|\mathsf{predicted} \cap \mathsf{oracle}|}
     {|\mathsf{oracle}|}
$$
where $\mathsf{predicted}$ is limited to the top 3 predictions of the
classier, and $\mathsf{oracle}$ is limited to the expressions that
changed \emph{and} were part of the type error slice. We make the
latter distinction as (1) changes that are not part of the type error
slice are noise in the data set, and (2) it makes the comparison between
feature sets easier to interpret since $\mathsf{oracle}$ never changes.
% NOTE: keep this at the end of the para or it screws up spacing...
\lstMakeShortInline{|}
%
\begin{table}[ht]
  \caption{
    Impact of Type Error Slice on Accuracy.
    \ES{TODO: load these numbers from CSV}
  }\label{tab:type-error-slice}
  \centering
  \begin{tabular}{lrcrrrrcrrrr}
    \toprule
                       &             & & \multicolumn{4}{c} \linear        & & \multicolumn{4}{c} \hiddenFH      \\
                                         \cmidrule{4-7}                        \cmidrule{9-12}
    Feature Set        & \# Features & & Top-1  & Top-2  & Top-3  & Recall & & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax       & 44          & & 23.5\% & 42.5\% & 55.9\% & 36.5\% & & 32.4\% & 49.3\% & 60.4\% & 39.6\% \\
    + \InSlice         & 45          & & 46.1\% & 64.9\% & 75.4\% & 48.8\% & & 56.1\% & 71.6\% & 82.7\% & 57.4\% \\
    Filter by \InSlice & 44          & & 54.9\% & 71.4\% & 82.5\% & 57.6\% & & 57.8\% & 72.8\% & 83.3\% & 57.8\% \\
    \bottomrule
  \end{tabular}
\end{table}

\autoref{tab:type-error-slice} shows the results of our experiment.
%
As expected, the baseline performs the worst, with a mere 23.6\% \linear
Top-1 accuracy.
%
Adding \InSlice improves the results substantially with a 46.4\% \linear Top-1
accuracy, demonstrating the importance of a minimal error slice.
%
However, filtering out expressions that are not part of the slice
\emph{further} improves the results to 54.9\% \linear Top-1 accuracy.
%
Interestingly, while the \hiddenFH performs similarly poor with no error
slice features, it recovers nearly all of its accuracy after being given
the error slice features.
%
Top-1 accuracy jumps from 32.4\% to 56.1\% when we add \InSlice, and only
improves by 1.7\% when we filter out expressions that are not part of the
error slice.

@@

We investigate the relative impact of the other
three classes of features discussed in \autoref{sec:features}, assuming
we have discarded expressions not in the type error slice.
%
For this experiment we consider again a baseline of only local syntactic
features, extended by each combination of
%
(1) expression size;
(2) contextual syntactic features; and
(3) typing features.
%
As before, we perform a 10-fold cross-validation with $\eta = 0.001$,
$\lambda = 0.001$, and a mini-batch size of 200, but we
train for a full 20 epochs to make the differences more apparent.
%
\begin{table}[ht]
  \caption{
    Impact of Contextual Features on Accuracy.
    \ES{TODO: load these numbers from CSV}
  }\label{tab:contextual-features}
  \centering
  \begin{tabular}{lrcrrrrcrrrr}
    \toprule
                             &             & & \multicolumn{4}{c} \linear        & & \multicolumn{4}{c} \hiddenFH      \\
                                               \cmidrule{4-7}                        \cmidrule{9-12}
    Feature Set              & \# Features & & Top-1  & Top-2  & Top-3  & Recall & & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax             &  44         & & 55.0\% & 71.6\% & 82.6\% & 57.7\% & & 57.7\% & 72.4\% & 83.0\% & 57.7\% \\
    \midrule
    + Size                   &  45         & & 55.8\% & 72.8\% & 82.5\% & 57.3\% & & 60.6\% & 75.0\% & 83.5\% & 57.2\% \\
    + Context                & 220         & & 59.9\% & 77.7\% & 86.4\% & 63.0\% & & 71.5\% & 84.4\% & 90.5\% & 68.8\% \\
    + Types                  & 102         & & 62.2\% & 77.7\% & 85.7\% & 62.2\% & & 74.0\% & 85.5\% & 90.8\% & 69.1\% \\
    \midrule
    + Context + Size         & 221         & & 60.2\% & 77.9\% & 86.0\% & 62.5\% & & 71.4\% & 84.1\% & 90.8\% & 69.0\% \\
    + Types + Size           & 103         & & 62.0\% & 78.2\% & 85.6\% & 62.3\% & & 73.7\% & 86.3\% & 90.9\% & 69.5\% \\
    + Context + Types        & 275         & & 63.2\% & 80.3\% & 87.9\% & 65.4\% & & 77.6\% & 88.1\% & 92.9\% & 72.9\% \\
    \midrule
    + Context + Types + Size & 276         & & 62.3\% & 80.0\% & 88.0\% & 65.4\% & & 76.8\% & 87.9\% & 92.7\% & 72.5\% \\
    \bottomrule
  \end{tabular}
  % \begin{minipage}{0.49\linewidth}
  % \centering
  % \hiddenF
  % \begin{tabular}{lrrrr}
  %   \toprule
  %   Feature Set                 & Top-1  & Top-2  & Top-3  & Recall \\
  %   \midrule
  %   Local Syntax                & 56.9\% & 72.2\% & 82.8\% & 57.9\% \\
  %   \midrule
  %   + Size                      & 59.7\% & 74.6\% & 83.0\% & 57.4\% \\
  %   + Context                   & 70.9\% & 83.7\% & 90.4\% & 69.2\% \\
  %   + Types                     & 72.1\% & 84.1\% & 90.3\% & 69.3\% \\
  %   \midrule
  %   + Size + Context            & 69.8\% & 83.5\% & 90.2\% & 68.6\% \\
  %   + Size + Types              & 72.3\% & 84.6\% & 90.3\% & 69.5\% \\
  %   + Context + Types           & 75.5\% & 86.4\% & 91.5\% & 71.7\% \\
  %   \midrule
  %   + All                       & 75.0\% & 86.8\% & 91.9\% & 72.0\% \\
  %   \bottomrule
  % \end{tabular}
  % \end{minipage}
\end{table}
%
\paragraph{Results}
\autoref{tab:contextual-features} summarizes the results of this experiment.
%
The \linear classifier and the \hiddenFH start off
competitive when given only local syntactic features, but the \hiddenFH
quickly outperforms as more features are added.

\ExprSize is the weakest feature, improving \linear Top-1
accuracy by only 1\% and \hiddenFH by only 4\%.
%
In contrast, the contextual syntactic features improve \linear Top-1
accuracy by 5\% (\resp 15\%), and the typing features improve
Top-1 accuracy by 7\% (\resp 17\%).
%
Furthermore, while \ExprSize does provide some benefit when it is the
only additional feature, it does not appear to provide any real increase
in accuracy when added alongside the contextual or typing features.
FIXME: Indicate that this is expected because of feature overlap: the
contextual ``child'' features partially encode subtree size.

As one might expect, the typing features are more beneficial than the
contextual syntactic features.
%
They improve Top-1 accuracy by an additional 2\%, and are much more
compact --- requiring only have 55 typing features compared to 176
contextual syntactic features.
%
This aligns with our intuition that types should be a good summary of
the context of an expression.
%
However, typing features do not \emph{subsume} contextual syntactic
features, we can gain an additional 1\% Top-1 accuracy (\resp 4\%) by
adding \emph{both}.

>>>>

We tested our hypothesis with the \linear and
%
\hiddenFH\footnote{A layer of 500 neurons is excessive when we have so few
  input features --- we use \hiddenFH for continuity with the
  surrounding sections.}
%
classifiers, cross-validated ($k=10$) over the combined SP14/FA15
dataset. We used a learning rate $\eta=0.001$, $L_2$ regularization rate
$\lambda=0.001$, and mini-batch size of 200. We trained for a single
epoch on feature sets (1) and (2), and for 8 epochs on (3), so that the
total number of training samples would be roughly equal for each feature
set.
%
\lstDeleteShortInline{|} % sigh...
In addition to the accuracy metric from before, we report each
classifier's \emph{recall} --- \ie ``How many true changes can we
remember?'' --- defined as
$$
\frac{|\mathsf{predicted} \cap \mathsf{oracle}|}
     {|\mathsf{oracle}|}
$$
where $\mathsf{predicted}$ is limited to the top 3 predictions of the
classier, and $\mathsf{oracle}$ is limited to the expressions that
changed \emph{and} were part of the type error slice. We make the
latter distinction as (1) changes that are not part of the type error
slice are noise in the data set, and (2) it makes the comparison between
feature sets easier to interpret since $\mathsf{oracle}$ never changes.
% NOTE: keep this at the end of the para or it screws up spacing...
\lstMakeShortInline{|}
%
\begin{table}[ht]
  \caption{
    Impact of Type Error Slice on Accuracy.
    \ES{TODO: load these numbers from CSV}
  }\label{tab:type-error-slice}
  \centering
  \begin{tabular}{lrcrrrrcrrrr}
    \toprule
                       &             & & \multicolumn{4}{c} \linear        & & \multicolumn{4}{c} \hiddenFH      \\
                                         \cmidrule{4-7}                        \cmidrule{9-12}
    Feature Set        & \# Features & & Top-1  & Top-2  & Top-3  & Recall & & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax       & 44          & & 27.7\% & 46.7\% & 58.6\% & 38.2\% & & 32.6\% & 49.5\% & 60.2\% & 39.4\% \\
    + \InSlice         & 45          & & 46.4\% & 65.1\% & 76.1\% & 48.9\% & & 55.4\% & 71.0\% & 82.1\% & 57.1\% \\
    Filter by \InSlice & 44          & & 55.9\% & 71.9\% & 82.9\% & 57.5\% & & 57.8\% & 72.7\% & 82.9\% & 57.6\% \\
    \bottomrule
  \end{tabular}
\end{table}

\autoref{tab:type-error-slice} shows the results of our experiment.
%
As expected, the baseline performs the worst, with a mere 27.7\% \linear
Top-1 accuracy.
%
Adding \InSlice improves the results substantially with a 46.4\% \linear Top-1
accuracy, demonstrating the importance of a minimal error slice.
%
However, filtering out expressions that are not part of the slice
\emph{further} improves the results to 55.9\% \linear Top-1 accuracy.
%
Interestingly, while the \hiddenFH performs similarly poor with no error
slice features, it recovers nearly all of its accuracy after being given
the error slice features.
%
Top-1 accuracy jumps from 32.4\% to 55.4\% when we add \InSlice, and only
improves by 2\% when we filter out expressions that are not part of the
error slice.

@@

We investigate the relative impact of the other
three classes of features discussed in \autoref{sec:features}, assuming
we have discarded expressions not in the type error slice.
%
For this experiment we consider again a baseline of only local syntactic
features, extended by each combination of
%
(1) expression size;
(2) contextual syntactic features; and
(3) typing features.
%
As before, we perform a 10-fold cross-validation with $\eta = 0.001$,
$\lambda = 0.001$, and a mini-batch size of 200, but we
train for a full 20 epochs to make the differences more apparent.
%
\begin{table}[ht]
  \caption{
    Impact of Contextual Features on Accuracy.
    \ES{TODO: load these numbers from CSV}
  }\label{tab:contextual-features}
  \centering
  \begin{tabular}{lrcrrrrcrrrr}
    \toprule
                             &             & & \multicolumn{4}{c} \linear        & & \multicolumn{4}{c} \hiddenFH      \\
                                               \cmidrule{4-7}                        \cmidrule{9-12}
    Feature Set              & \# Features & & Top-1  & Top-2  & Top-3  & Recall & & Top-1  & Top-2  & Top-3  & Recall \\
    \midrule
    Local Syntax             &  44         & & 56.6\% & 72.6\% & 83.2\% & 57.7\% & & 57.6\% & 72.6\% & 83.1\% & 57.7\% \\
    \midrule
    + Size                   &  45         & & 57.2\% & 73.7\% & 83.0\% & 57.4\% & & 60.9\% & 75.5\% & 84.1\% & 57.9\% \\
    + Context                & 220         & & 61.1\% & 78.7\% & 86.7\% & 63.0\% & & 71.5\% & 84.8\% & 90.8\% & 69.0\% \\
    + Types                  & 102         & & 63.1\% & 77.8\% & 85.2\% & 61.6\% & & 73.4\% & 85.4\% & 90.7\% & 68.9\% \\
    \midrule
    + Context + Size         & 221         & & 61.1\% & 78.8\% & 86.3\% & 62.4\% & & 71.5\% & 84.4\% & 90.8\% & 69.0\% \\
    + Types + Size           & 103         & & 61.9\% & 78.7\% & 85.6\% & 62.0\% & & 73.1\% & 85.2\% & 91.1\% & 69.5\% \\
    + Context + Types        & 275         & & 63.1\% & 80.9\% & 88.0\% & 65.0\% & & 77.2\% & 88.3\% & 92.5\% & 72.6\% \\
    \midrule
    + Context + Types + Size & 276         & & 62.8\% & 80.5\% & 88.1\% & 65.1\% & & 77.3\% & 88.1\% & 92.7\% & 72.7\% \\
    \bottomrule
  \end{tabular}
  % \begin{minipage}{0.49\linewidth}
  % \centering
  % \hiddenF
  % \begin{tabular}{lrrrr}
  %   \toprule
  %   Feature Set                 & Top-1  & Top-2  & Top-3  & Recall \\
  %   \midrule
  %   Local Syntax                & 56.9\% & 72.2\% & 82.8\% & 57.9\% \\
  %   \midrule
  %   + Size                      & 59.7\% & 74.6\% & 83.0\% & 57.4\% \\
  %   + Context                   & 70.9\% & 83.7\% & 90.4\% & 69.2\% \\
  %   + Types                     & 72.1\% & 84.1\% & 90.3\% & 69.3\% \\
  %   \midrule
  %   + Size + Context            & 69.8\% & 83.5\% & 90.2\% & 68.6\% \\
  %   + Size + Types              & 72.3\% & 84.6\% & 90.3\% & 69.5\% \\
  %   + Context + Types           & 75.5\% & 86.4\% & 91.5\% & 71.7\% \\
  %   \midrule
  %   + All                       & 75.0\% & 86.8\% & 91.9\% & 72.0\% \\
  %   \bottomrule
  % \end{tabular}
  % \end{minipage}
\end{table}
%
\paragraph{Results}
\autoref{tab:contextual-features} summarizes the results of this experiment.
%
The \linear classifier and the \hiddenFH start off
competitive when given only local syntactic features, but the \hiddenFH
quickly outperforms as more features are added.

\ExprSize is the weakest feature, improving \linear Top-1
accuracy by less than 1\% and \hiddenFH by only 3\%.
%
In contrast, the contextual syntactic features improve \linear Top-1
accuracy by 5\% (\resp 14\%), and the typing features improve
Top-1 accuracy by 7\% (\resp 16\%).
%
Furthermore, while \ExprSize does provide some benefit when it is the
only additional feature, it does not appear to provide any real increase
in accuracy when added alongside the contextual or typing features.
\ES{FIXME: Indicate that this is expected because of feature overlap: the
contextual ``child'' features partially encode subtree size.}

As one might expect, the typing features are more beneficial than the
contextual syntactic features.
%
They improve Top-1 accuracy by an additional 2\%, and are much more
compact --- requiring only have 55 typing features compared to 176
contextual syntactic features.
%
This aligns with our intuition that types should be a good summary of
the context of an expression.
%
However, typing features do not appear to \emph{subsume} contextual
syntactic features, the \hiddenFH gains an additional 4\% Top-1 accuracy
when both are added.

### 2017-04-07 paper/oopsla17-submission/evaluation.tex
update numbers

\paragraph{Feature Selection}
We extract a set of 277 features from each sub-expression in a student
program, including:
%
\begin{enumerate}
\item 44 local syntactic features. In addition to the syntax of \lang,
  we support the full range of arithmetic operators (integer and
  floating point), equality and comparison operators, character and
  string literals, and a user-defined |expr| type of simple arithmetic
  expressions. We discuss the challenge of supporting other
  user-defined types in \autoref{sec:discussion}.
\item 176 contextual syntactic features. For each sub-expression we
  additionally extract the local syntactic features of its parent and
  first, second, and third (left-to-right) children. If an expression
  does not have a parent or children, these features will simply be
  disabled. If an expression has more than three children, the
  classifiers will receive no information about the additional
  children.
\item 55 typing features. In addition to the types of \lang, we support
  |int|s, |float|s, |char|s, |string|s, and the user-defined |expr|
  mentioned above. These features are extracted for each sub-expression
  and for the contextual sub-expressions.
\item one feature denoting the size of each sub-expression.
\item one feature denoting whether each sub-expression is part of the
  minimal type error slice. We use this feature as a ``hard''
  constraint, sub-expressions that are not part of the minimal slice
  will be preemptively discarded, we justify this decision in
  \autoref{sec:feature-utility}.
\end{enumerate}

>>>>

\paragraph{Feature Selection}
We extract a set of 277 features from each sub-expression in a student
program, including:
%
\begin{enumerate}
\item 44 local syntactic features. In addition to the syntax of \lang,
  we support the full range of arithmetic operators (integer and
  floating point), equality and comparison operators, character and
  string literals, and a user-defined |expr| type of simple arithmetic
  expressions. We discuss the challenge of supporting other
  user-defined types in \autoref{sec:discussion}.
\item 176 contextual syntactic features. For each sub-expression we
  additionally extract the local syntactic features of its parent and
  first, second, and third (left-to-right) children. If an expression
  does not have a parent or children, these features will simply be
  disabled. If an expression has more than three children, the
  classifiers will receive no information about the additional
  children.
\item 55 typing features. In addition to the types of \lang, we support
  |int|s, |float|s, |char|s, |string|s, and the user-defined |expr|
  mentioned above. These features are extracted for each sub-expression
  and for the contextual sub-expressions.
\item One feature denoting the size of each sub-expression.
\item One feature denoting whether each sub-expression is part of the
  minimal type error slice. We use this feature as a ``hard''
  constraint, sub-expressions that are not part of the minimal slice
  will be preemptively discarded, we justify this decision in
  \autoref{sec:feature-utility}.
\end{enumerate}
